---
title: "Introduction to R"
author: "Revendranath T"
date: "2/25/2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Using R Console

 1+1
- max(1,2,5)
- q() #exit from R & RStudio
- escape key to interrupt R
- use help() or ? to get help from R
- example help(mean or ?mean)

## Slide with R Output

```{r basics, echo = TRUE}
pi
sqrt(2)
5^2
```

## Use print
- To format any R value for printing, including structured values such as matrices and lists
```{r print, echo=false}
#print matrix
print(matrix(c(1, 2, 3, 4), 2, 2))

#print list
print(list("a", "b", "c"))

```


## Working with variables
- No need to declare variables first = or <-
- Global variables: <<-
- List variables: ls() or ls.str()
- Delete variables: rm()

```{r variables, echo=TRUE}
# = or <- are used for assignment
x = 3
y <- 4


x = c("fee", "fie", "foe", "fum") # x is a vector
print(x)

#global variable assignment <<-
#That forces the assignment to a global variable rather than a local variable
x <<- 3

#listing variables
ls()
ls.str()

# delete a variable
rm(x)

# delete multiple variables
rm(x,y)
```

### Create vectors
- Use the c(...) operator to construct a vector from given values
- Use mode to know the type of vector: numeric, character, boolean

```{r vectors, echo=TRUE}
x1 = c(1, 1, 2, 3, 5, 8, 13, 21)
mode(x1)

f1 <- c(1 * pi, 2 * pi, 3 * pi, 4 * pi)
mode(f1)
s1 <- c("Welcome", "to", "BITS", "Pilani", "!")
mode(s1)

l1 <- c(TRUE, TRUE, FALSE, TRUE)
mode(l1)

```
### Modes
- What happens when you merge vectors of two different modes?
```{r modes, echo=TRUE}

v1 <- c(1, 2, 3)
mode(v1)
v3 <- c("A", "B", "C")
mode(v3)
v2 <- c(v1, v3)
mode(v2)
```

## Basic Statistics with R
- mean()
- median()
- var()
- sd()
- cor()
- cov()

```{r, echo=TRUE}
#assign values to variable x
x <- c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34)

#mean
mean(x)

#median
median(x)

#variance
var(x)

#standard deviation
sd(x)

#correlation & covariance. Need 2 variables. 

# create a y variable
y <- log(x + 1)

#correlation
cor(x, y)
#covariance
cov(x, y)

```

## Handling null values or NA
- exclude null values using na.rm = TRUE
- use imputation methods (discussed later)


```{r, echo=TRUE}
x <- c(0, 1, 1, 2, 3, NA)

#without na.rm = TRUE => Excel would have required delete/handling missing value
mean(x)
sd(x)

#with na.rm = TRUE
mean(x,na.rm = TRUE) #ignoring null value
sd(x,na.rm = TRUE)

```

## Basic statistics for multiple columns
- use map_dbl() function from purrr package in tidyverse

```{r, echo=TRUE}
# use ?mtcars to verify the dataset
library(tidyverse)
data(mtcars)
str(mtcars)
head(mtcars)
#mean
map_dbl(mtcars, mean)

#median
map_dbl(mtcars, median)

#sd
map_dbl(mtcars, sd)

#var
map_dbl(mtcars, var)

#all summary statistics
summary(mtcars)

#correlation
cor(mtcars)

#covariance
cov(mtcars)
```


### Create sequences
- create a sequence of numbers
- Use an n:m expression to create the simple sequence n, n+1, n+2, …, m
- Use the `seq` function for sequences with an increment other than 1
- Use the rep function to create a series of repeated values

```{r sequences, echo=TRUE}
1:5

seq(from = 1, to = 20, by = 2)

rep(8, times = 5)
```

## Comparing vectors
```{r, echo=TRUE}
a <- 3
a

a == pi # Test for equality

a != pi # Test for inequality

a < pi

a > pi

a <= pi

a >= pi

```


```{r, echo=TRUE}
v <- c(3, pi, pi, 0)

w <- c(pi, pi)

v == w # Compare two vectors with different sizes

v
v < 3.14

v != w

v < w

v <= w

v > w

v >= w

```

```{r, echo=TRUE}
v <- c(3, pi, 4)
v == pi # Compare a 3-element vector against one number

v != pi

```

## Select elements in the vector
```{r, echo=TRUE}
fib <- c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34)
fib

fib[1]

fib[2]

fib[3]

fib[4]

fib[5]

```

```{r, echo=TRUE}
fib[1:3] # Select elements 1 through 3

fib[4:9] # Select elements 4 through 9

fib[c(1, 2, 4, 8)]

fib[-1] # Ignore first element

fib[1:3] # As before

fib[-(1:3)] # Invert sign of index to exclude instead of select

```


```{r, echo=TRUE}
fib < 10 # This vector is TRUE wherever fib is less than 10

fib[fib < 10] # Use that vector to select elements less than 10

fib %% 2 == 0 # This vector is TRUE wherever fib is even

fib[fib %% 2 == 0] # Use that vector to select the even elements
```


## Combining vector comparisons, logical operators, and vector indexing

```{r, echo=TRUE}
v <- c(3, 6, 1, 9, 11, 16, 0, 3, 1, 45, 2, 8, 9, 6, -4)

v[ v > median(v)]

v[ (v < quantile(v, 0.05)) | (v > quantile(v, 0.95)) ]

#Select all elements that exceed ±1 standard deviations from the mean:

v[ abs(v - mean(v)) > sd(v)]

#Select all elements that are neither NA nor NULL:

v <- c(1, 2, 3, NA, 5)
v[!is.na(v) & !is.null(v)]

```

## Vector Arithmetic
```{r, echo=TRUE}
v <- c(11, 12, 13, 14, 15)
w <- c(1, 2, 3, 4, 5)
v + w

v - w

v * w

v / w

w^v

```


## vector-scalar operations
```{r, echo=TRUE}
w

w + 2

w - 2

w * 2

w / 2

2^w
```

```{r, echo=TRUE}
w

#sqrt
sqrt(w)

#log
log(w)

#sin
sin(w)

#mean
mean(w)

#distance from mean
w - mean(w)

#sd
sd(w)

#z-score
(w - mean(w))/sd(w)


```

##  Choosing Between Two Alternatives: if-else

```{r, echo=TRUE}
#syntax 
#if (condition) {
  ##do this if condition is TRUE 
#} 
#else {
  ## do this if condition is FALSE
#}


#Example

if (x >= 0) {
  print(sqrt(x))             # do this if x >= 0
} else {
  print("negative number")   # do this otherwise
}

x <- c(-2, -1, 0, 1, 2)

if (all(x < 0)) {
  print("all are negative")
}

if (any(x < 0)) {
  print("some are negative")
}
```


##  Iterating with a Loop
```{r, echo=TRUE}
#Syntax
for (x in v) {
  # do something with x
}

for (x in 1:5) {
  #cat(x, x^2, "\n")
  print(x) 
  print( x^2)
}


v <- 1:5
for (i in 1:5) {
  v[[i]] <- v[[i]] ^ 2
}
print(v)

```


## functions
```{r, echo=TRUE}
#syntax
name <- function(param1, ..., paramN) {
          expr1
          .
          .
          .
          exprM
}

#function for coefficient of variation
cv <- function(x) {
  sd(x) / mean(x)
}

cv(1:10)     # Set x = 1:10 and evaluate sd(x)/mean(x)
```

